---
title: "Getting Started with haze"
author: "Tim Schäfer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with haze}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`haze` provides fast smoothing of per-vertex data on triangular meshes, along with utility functions for mesh manipulation and data interpolation. The smoothing is implemented in C++ and supports parallel processing for multiple overlays.

## Smoothing Per-Vertex Data

The main workhorse is `pervertexdata.smoothnn()`, which performs nearest-neighbor smoothing using the $k$-ring neighborhood of each vertex.

```{r}
library("haze")

# Load a FreeSurfer mesh and per-vertex data
mesh <- freesurferformats::read.fs.surface(
  system.file("extdata", "fsaverage_mesh_lh_white", package = "haze", mustWork = TRUE))
thickness <- freesurferformats::read.fs.morph(
  system.file("extdata", "fsaverage_lh_thickness", package = "haze", mustWork = TRUE))

# Smooth: k=2 neighborhood, 300 iterations
smoothed <- pervertexdata.smoothnn(mesh, thickness, num_iter = 300L, k = 2L)
```

## Re-Using the Mesh Neighborhood

When smoothing many datasets on the same mesh, pre-compute the adjacency list once with `mesh.adj()` and then use `pervertexdata.smoothnn.adj()`:

```{r}
mesh_adj <- mesh.adj(mesh, k = 1L)
```

This adjacency can be reused:

```{r}
data1 <- rnorm(length(mesh_adj), mean = 1.0, sd = 0.1)
data2 <- rnorm(length(mesh_adj), mean = 5.0, sd = 0.1)

smoothed1 <- pervertexdata.smoothnn.adj(mesh_adj, data1, num_iter = 15L)
smoothed2 <- pervertexdata.smoothnn.adj(mesh_adj, data2, num_iter = 15L)
```

Multiple overlays (as a matrix, one per row) are automatically smoothed in parallel:

```{r}
pvd <- rbind(data1, data2)
options("mc.cores" = 2L)
smoothed_pvd <- pervertexdata.smoothnn.adj(mesh_adj, pvd, num_iter = 15L)
```

## Creating Sub-Meshes

Use `submesh.vertex()` to extract a patch of a mesh defined by vertex indices:

```{r}
# Extract a sub-mesh of the first 1000 vertices
small_mesh <- submesh.vertex(mesh, vertex_indices = 1:1000)
```

## Working with k-d Trees

`find_nv_kdtree()` finds the nearest mesh vertex for given query coordinates:

```{r}
query_points <- matrix(c(0, 0, 0, 10, 10, 10), ncol = 3, byrow = TRUE)
nearest_vertex_idx <- find_nv_kdtree(query_points, mesh$vb[1:3, ])
```

`nn_interpolate_kdtree()` assigns per-vertex data to query points via nearest-neighbor lookup, while `linear_interpolate_kdtree()` performs linear interpolation — useful for mapping morphometry data between subjects with aligned spherical meshes.

## Mesh Conversion Utilities

Convert between FreeSurfer surfaces and `rgl` `tmesh3d` objects:

```{r}
tmesh <- fs.surface.to.tmesh3d(mesh)
fs_surface <- tmesh3d.to.fs.surface(tmesh)
```

## Further Reading

- Full API reference: [dfsp-spirit.github.io/haze/reference/](https://dfsp-spirit.github.io/haze/reference/)
- Run `example(<function>)` in R for interactive demos of any function.
